---
title: "NMBU PARAFAC_creating the model"
output: pdf_document
date: "2024-12-03"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Creating a PARAFAC model
The data has now been uploaded and preprocessed, and indices and Coble peaks been calculated. 
Note that from the preprocessing, from inner filter correction, it appeared that several of the samples should have been diluted prior to analysis. See warning messages. Only "unten" samples. 
```{r Loading required packages required packages, message=FALSE, warning=FALSE}
Packages <- c("staRdom", "dplyr", "tidyr", "data.table", "reshape")
#"libwgeom",
lapply(Packages, library, character.only = TRUE)
cores <- detectCores(logical = FALSE)
```

## 1) Looking for noise 
```{r looking for noise}
# Restore the object
eemint4 <- readRDS(file = "NMBU_PARAFAC_datset.rds")
eem_overview_plot(eemint4[1:9], spp=9, contour = TRUE)
```
Taking a visual inspection of all sample EEMS. 
Several samples looks strange at around ex/em 400/400 nm, for example sample untenGF117 in figure 
This region should be removed and the signal interpolated. 

Plotting all samples with such strange signal.

Tried to remove and interpolate the region, but did not manage... This region has been found to influence the model components so should be removed. Consider also other regions. 

```{r remove noise at around 400/400}
#taking a look at those samples with 
eemint4 %>% 
  eem_extract(sample = c("untenGF97", "untenGF99", "untenGF88", "untenGF69", "untenGF68", "untenGF48", "untenGF317", "untenGF297", "untenGF288", "untenGF28", "untenGF277", "untenGF268", "untenGF257", "untenGF239", "untenGF237", "untenGF217", "untenGF198", "untenGF168", "untenGF167", "untenGF148", "untenGF139", "untenGF128", "untenGF114", "untenGF117",
                         "obenGF317", "obenGF297", "obenGF197", "obenGF167"), keep = TRUE) %>%
  ggeem(contour = TRUE)


eem_list <- eemint4 %>%
  eem_setNA(sample = c("untenGF97", "untenGF99", "untenGF88", "untenGF69", "untenGF68", "untenGF48", "untenGF317", "untenGF297", "untenGF288", "untenGF28", "untenGF277", "untenGF268", "untenGF257", "untenGF239", "untenGF237", "untenGF217", "untenGF198", "untenGF168", "untenGF167", "untenGF148", "untenGF139", "untenGF128", "untenGF114", "untenGF117",
                         "obenGF317", "obenGF297", "obenGF197", "obenGF167"), ex = 375:430, em= 400:450, interpolate = TRUE) %>%
  eem_setNA(em= 400:450, interpolate = TRUE)

eem_list %>% 
  eem_extract(sample = c("untenGF97", "untenGF99", "untenGF88", "untenGF69", "untenGF68", "untenGF48", "untenGF317", "untenGF297", "untenGF288", "untenGF28", "untenGF277", "untenGF268", "untenGF257", "untenGF239", "untenGF237", "untenGF217", "untenGF198", "untenGF168", "untenGF167", "untenGF148", "untenGF139", "untenGF128", "untenGF114", "untenGF117",
                         "obenGF317", "obenGF297", "obenGF197", "obenGF167"), keep = TRUE) %>%
  ggeem(contour = TRUE) 

```
 
## 1) First attempt PARAFAC model - quick
First attempt to find approximate number of components before a more refined model is created. 
```{r}
dim_min <- 3
dim_max <- 7

nstart <- 25 # number of similar models from which best is chosen
maxit = 5000 # maximum number of iterations in PARAFAC analysis
ctol <- 10^-6 # tolerance in PARAFAC analysis
```

Create PARAFAC moodel with non-negative constraint. 
```{r}
pf1 <- eem_parafac(eem_list, comps = seq(dim_min,dim_max), normalise = FALSE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)
eempf_compare(pf1, contour = TRUE)
```

From the model fit: 4 components or more
Visual inspection of the components is easier in the normalised version next
# rescale B and C modes to a maximum fluorescence of 1 for each component
```{r}
pf1n <- lapply(pf1, eempf_rescaleBC, newscale = "Fmax")
eempf_compare(pf1n, contour = TRUE)
```

- from the excitation and emission spectra, there appear to be some noise? 
- looks like there might be some strange signal in the 400/400 region for some of the components. Perhaps worth going back to remove that region. 
- several aspects with the excitation and emission spectral lines that do not make sense. May need to go back to clean up. Or may improve with a more refined model. 

## 2) check for correlation between the components
- there appear to be correlation between the components so selects normalisation
```{r check for correlation}
eempf_cortable(pf1n[[3]], normalisation = FALSE)
eempf_corplot(pf1n[[3]], progress = FALSE, normalisation = FALSE)
```
- high correlation between some of the components, normalisation should be done

Normalise to reduce correlation between components
wide range of DOC among samples normally require this step
```{r normalise}
pf2 <- eem_parafac(eem_list, comps = seq(dim_min,dim_max), normalise = TRUE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)
# rescale B and C modes
pf2nx <- lapply(pf2, eempf_rescaleBC, newscale = "Fmax")
eempf_compare(pf2nx, contour = TRUE, normalise=FALSE)
```

#Set normalisation TRUE if you want to see the actual data
```{r}
eempf_cortable(pf2nx[[3]], normalisation = FALSE)
eempf_corplot(pf2nx[[3]], progress = FALSE, normalisation = FALSE)
#eempf_compare(pf2nx, contour = TRUE, normalisation=FALSE)
#eempf_plot_comps(pf2, contour = TRUE, type = 1)
```

## 3) Find and exclude outlier leverages
- plot leverage (nice plot)
- plot leverage, not so nice plot but interactive to select what to exclude
- saved in exclude, can be used to start over again with eem_list_ex <- eem_list %>% eem_exclude(exclude) above
- NOt all potential outliers from leverage plot are potential outliers from the other plot. 
```{r calculate leverage}
cpl <- eempf_leverage(pf2nx[[3]])
#cpl <- eempf_leverage(pf3n[[2]])

eempf_leverage_plot(cpl,qlabel=0.1)
#exclude <- eempf_leverage_ident(cpl,qlabel=0.1)

```

should remove sample obenGF317!
additional samples or wavelengths that should be removed?

```{r r PLot leverage of potential outliers}
eempf_residuals_plot(pf2nx[[3]], eem_list, residuals_only = TRUE, select = c("obenGF317"), spp = 6, cores = cores, contour = TRUE)

# samples, excitation and emission wavelengths to exclude, makes sense after calculation of leverage
exclude <- list("ex" = c(),
                "em" = c(),
                "sample" = c("obenGF317")
)

# exclude outliers if neccessary. if so, restart analysis
eem_list_ex <- eem_exclude(eem_list, exclude)
```

## 4) Remake the model - quick
```{r}
pf3 <- eem_parafac(eem_list_ex, comps = seq(dim_min,dim_max), normalise = TRUE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)
# rescale B and C modes
pf3nx <- lapply(pf3, eempf_rescaleBC, newscale = "Fmax")
eempf_compare(pf3nx, contour = TRUE, normalise=FALSE)

cpl <- eempf_leverage(pf3nx[[3]])
eempf_leverage_plot(cpl,qlabel=0.1)

#exclude <- eempf_leverage_ident(cpl,qlabel=0.1) #sample obenGF164?
eempf_residuals_plot(pf3nx[[3]], eem_list_ex, residuals_only = TRUE, select = c("obenGF164"), spp = 6, cores = cores, contour = TRUE)

```
decided to keep sample obenGF164 for mow, may need to revisit. 

## 5) Remake the PARAFAC MODEL - refined
- likely need for more cleanup? 
- the excitation and emission lines does not look right.. 
```{r Re-make PARAFAC model}
ctol <- 10^-8     # decrease tolerance in PARAFAC analysis
nstart = 20        # increase number of random starts
maxit = 10000      # increase number of maximum interations

pf4fc <- eem_parafac(eem_list_ex, comps = 5, normalise = TRUE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, cores = cores, output="all")
pf4fnc <- lapply(pf4fc, eempf_rescaleBC, newscale = "Fmax")
eempf_compare(pf4fnc, contour = TRUE, normalise=FALSE)

summary(pf4fnc)
#something on the importance of each component
b <- eempf_varimp(pf4fnc[[1]], eem_list_ex, cores = cores)
b

#eempf_comp_load_plot(pf4fnc[[1]], contour = TRUE)
ggeem(pf4fnc[[1]], contour = TRUE)
```

If some splits look similar (or see below, have a Tucker’s congruency coefficient close to 1) and others do not, this is a strong hint that the particular splits or the values for maxit and nstart might be responsible for unstable conditions. Problems can be caused by splitting the data according to natural differences (e.g. sampling date A and sampling date B) which will then lead to different PARAFAC models. 

### 3.1) Model validation
There are different ways of checking the model. With natural DOM samples, it is very unlikely to "pass" all of the tests. An overall evaluation should be made. Also, each test has their pros and cons. 

#### Leverages
```{r CHeking and plotting the model}
#leverages, outliers?
cplx <- eempf_leverage(pf4fnc[[1]])
eempf_leverage_plot(cplx,qlabel=0.1)

#residuals, if samples appear suspect from the leverage check
#eempf_residuals_plot(pf4fnc[[1]], eem_list7, residuals_only = TRUE, select = c("DOM11856","DOM12210"), spp = 6, #cores = cores, contour = TRUE)

#plot model
eempf_comp_load_plot(pf4fnc[[1]], contour = TRUE)

#Check the convergence behaviour of the created models:
eempf_convergence(pf4fnc[[1]])

#plot residuals 
eempf_residuals_plot(pf4fnc[[1]], eem_list_ex, select = eem_names(eem_list_ex), cores = cores, contour = TRUE) 

#Plot the resulting components and loadings
eempf_comp_load_plot(pf4fc[[1]], contour = TRUE)
#eempf_residuals_plot(pf4f[[1]], eem_list6, cores = cores, contour = TRUE)

#SPlit-half analysis
sh <- splithalf(eem_list_ex, 5, normalise = TRUE, rand = TRUE, cores = cores, nstart = nstart, maxit = maxit, ctol = ctol) #HER MÅ DET OGSÅ VURDERES!
splithalf_plot(sh)

#Tucker’s congruency coefficient 
tcc_sh_table <- splithalf_tcc(sh)
tcc_sh_table

#converging models
eempf_convergence(pf4fc[[1]])
```

#To create output file
```{r}
eempf_openfluor(pf4fc[[1]], file = "241204_NMBU_PARAFAC_initial test.txt")

```

